---
title: "Variance estimation: ANCOVA RCT analysis"
author: "Dominic Magirr"
date: "2025-10-05"
categories: [RCT]
---

Consider a 2-arm RCT (1:1 allocation) with a continuous outcome, total sample size 200 and seven baseline covariates. Consider a superpopulation set-up. The estimand is the (super) population average treatment effect $\theta = E(Y(1) - Y(0))$. Estimation is performed via a linear regression working model

$$E(Y_i \mid A_i, X_{1,i},ldots,X_{7,i}) = \beta_0 + \theta A_i + \beta_1 X_{1,i} + \cdots + \beta_7{X_{7,i}}$$

so that $\hat{\theta}$ is the usual least squares estimator of $\theta$. For example,

```{r}
set.seed(620)

n <- 200
p <- 7

## simulate covariates 
X_star <- matrix(rnorm(n * p), nrow = n, ncol = p)

## equal allocation
a <- rep(c(0, 1), each = n / 2)

## design matrix
X <- cbind(1, a, X_star)

## true parameter values
beta <- c(0, 0.2, rep(0.1, p))

## simulate outcome
y <- rnorm(n, X%*%beta)

## data set
dat <- as.data.frame(cbind(y, a, X_star))
dat$a <- as.factor(dat$a)

## fit model
fit <- lm(y ~ ., data = dat)

## point estimate theta_hat
fit$coef[2]
```

How should we estimate the variance of $\hat{\theta}$? One option is the usual ANCOVA approach,

```{r}
## Model based Var
var_model <- vcov(fit)["a1", "a1"]
var_model
```

Alternatively, we could use an influence function approach, as for example described in [Ye et al. (2023) "Toward Better Practice of Covariate Adjustment in Analyzing Randomized Clinical Trials"](<https://doi.org/10.1080/01621459.2022.2049278>), which has the endorsement of being included in [recent FDA guidance](<https://www.fda.gov/regulatory-information/search-fda-guidance-documents/adjusting-covariates-randomized-clinical-trials-drugs-and-biological-products>), and is implemented in [{RobinCar2}](<https://openpharma.github.io/RobinCar2/main/>),

```{r}
library(RobinCar2)
## RobinCar2 Var
robin_fit <- robin_lm(as.formula(paste0("y ~ ", paste0(names(dat)[-1], collapse = "+"))),
                      data = dat,
                      treatment = a ~ sp(1))

var_robin <- robin_fit$contrast$variance[1,1]
var_robin
```

This estimated variance is 15\% smaller than the usual ANCOVA variance estimate

```{r}
var_robin / var_model
```

Same model, same data, 15\% smaller estimated variance.

## So what?

The methods included in {RobinCar2} are powerful and useful. I'm an advocate for using more covariate adjustment in the primary analysis of RCTs. I'm especially excited about the methods for [covariate adjustment involving time-to-event outcomes](<https://academic.oup.com/biomet/article-pdf/111/2/691/57467434/asad045.pdf>). They need to be used in the appropriate settings, however.

It's easy to say that's when \$p\$ is not too large, and \$n\$ is not too small. I chose this example to be somewhere on the boundary of what I would instinctively consider reasonable but can still lead to a dramatic difference between the model-based and influence function approaches.

I'm slowly building an understanding of what's driving this difference. It's a combination of several factors: conditional vs unconditional inference, variance inflation factors [(see Senn et al.,2024)](<https://arxiv.org/pdf/2408.06760>), and a degrees of freedom correction. In large(ish) sample sizes each of these factors might not seem to make a huge difference in isolation, but together it could add up to a big difference.
