---
title: "Simulate from a Normal-inverse-Wishart distribution"
author: "Dominic Magirr"
date: "2023-08-08"
categories: [R]
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```



## The problem

 

- Given some data which I assume is multivariate Gaussian, I'd like to simulate from the posterior distribution of the model parameters.

 

- The conjugate Bayesian analysis is a [Normal-inverse-Wishart distribution](https://en.wikipedia.org/wiki/Normal-inverse-Wishart_distribution).

 

- I'd like to do this very quickly based on snippets from wikipedia, rather than thinking much about it.

 

## Snippets from wikipedia

 

Here is the first snippet...

 

> To sample from the joint posterior of $(\mu, \Sigma)$, one simply draws samples from $\Sigma \mid y \sim W^{-1}(\Psi_n, \nu_n)$, then draw $\mu \mid \Sigma, y \sim N_p(\mu_n, \Sigma / \lambda_n)$, where

$$\mu_n = \frac{\lambda\mu_0+n\bar{y}}{\lambda + n}$$

$$\lambda_n = \lambda + n$$

$$\nu_n = \nu + n$$

$$\Psi_n = \Psi + S + \frac{\lambda n}{\lambda + n}(\bar{y}-\mu_0)(\bar{y}-\mu_0)^T$$

where

$$S = \sum_{i=1}^n  (y_i -\bar{y})(y_i - \bar{y})^T.$$

 

 

Ok, so this assumes we have used a prior distribution:

 

$$\Sigma \sim W^{-1}(\Psi, \nu)$$

$$\mu \mid \Sigma \sim N(\mu_0, \Sigma/\lambda)$$

I don't want to think too hard about prior distributions, even though that's dangerous. It isn't the point of this exercise. I want $\mu_0$, $\lambda$, $\nu$ and $\Psi$ to be small. Let's just say zero for now, even if that's a terrible idea. This way I've reduced the problem to drawing samples from $\Sigma \mid y \sim W^{-1}(S, n)$, then $\mu \mid \Sigma, y \sim N_p(\bar{y}, \Sigma / n)$.

 

 

The second snippet [(from a different page)](https://en.wikipedia.org/wiki/Inverse-Wishart_distribution) is...

 

> If $X\sim W(\Sigma, \nu)$, then $A = X^{-1}$ has an inverse-Wishart distribution $A \sim W^{-1}(\Sigma^{-1}, \nu).$

 

Ok, so this is mixing up notation now. Our target in our original notation is $\Sigma \mid y \sim W^{-1}(S, n)$. This snippet says we can achieve that by simulating $X$ from a Wishart distribution $X\sim W(S^{-1}, n)$, and then taking $\Sigma = X^{-1}$.

 

## Example in R

 

```{r}

# simulate some 3-dimensional normal data, n = 20

n <- 20

Y <- mvtnorm::rmvnorm(n = n,

                      mean = c(180, 190, 200),

                      sigma = matrix(70, 3, 3) + diag(30, 3))

 

# calculate Y_bar and S

Y_bar <- colMeans(Y)

S <- (t(Y) - Y_bar) %*% t(t(Y) - Y_bar)

 

# simulate from inverse-Wishart distribution

X <- rWishart(1, df = n, Sigma = solve(S))

Sigma <- solve(X[,,1])

Sigma

 

## simulate from normal distribution

mu <- mvtnorm::rmvnorm(1, mean = Y_bar, sigma = Sigma / n)

mu

```

 